### 2014-11-26 turcksin-spring-mechanics/paper.tex
Add the first two pages.

\documentclass{article}
\usepackage{fullpage}
\usepackage{wrapfig}
\usepackage[colorlinks,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
\usepackage{amsmath,amsfonts}
\usepackage{graphicx,color}
\usepackage{listings}
\lstset{language=Python,basicstyle=\ttfamily\footnotesize}
\usepackage{verbatim}

\newcommand{\dealii}{{\textsc{deal.II}}}
\newcommand{\pfrst}{{\normalfont\textsc{p4est}}}
\newcommand{\trilinos}{{\textsc{Trilinos}}}
\newcommand{\aspect}{\textsc{Aspect}}
\newcommand{\petsc}{\textsc{PETSc}}

\newtheorem{remark}{Remark}

\setlength{\emergencystretch}{20pt}

\lstset{language=C++,%
  basicstyle=\small\ttfamily,%
  escapeinside={\%*}{*)},%
  keywordstyle={},%
  captionpos=b,%
  frame=single,basicstyle=\footnotesize%
}
\lstset{language=Python,%
  basicstyle=\small\ttfamily,%
  escapeinside={\%*}{*)},%
  keywordstyle={},%
  captionpos=b,%
  frame=single,basicstyle=\footnotesize%
}
%

\begin{document}

%\runningheads{B. Turcksin, T. Heister, W. Bangerth}{Clone and graft: Testing whole applications}

\title{Clone and graft:\\ Testing whole scientific
  applications as they are built}

\author{Bruno Turcksin\footnote{Department of Mathematics, Texas A\&M
    University, College Station, TX 77843-3368, USA, \url{???}}
\and
   Timo Heister\footnote{Mathematical Sciences, O-110 Martin Hall, Clemson University,
     Clemson, SC 29634-095, USA, \url{heister@clemson.edu}}
\and
   Wolfgang Bangerth\footnote{Corresponding author. Department of Mathematics, Texas A\&M
     University, College Station, TX 77843-3368, USA, \url{bangerth@math.tamu.edu}}}

\maketitle

\section{Introduction}

Our group has been building scientific software for more than a decade and a
half by now. Over the years we have developed many procedures that help us
test our software extensively but our expertise was mostly in the development
of \textit{libraries} for numerical methods -- specifically, the \dealii{}
library for finite element computations (see
\url{http://www.dealii.org/}). Libraries are of course the foundation of
almost every single scientific code, starting with ``simple'' ones such as
BLAS and LAPACK to very much more complex ones such as \petsc{}, \trilinos{}, or
\dealii{}.

Developing testing schemes for libraries is reasonably well understood. So,
when we ventured in the area of building applications for scientific purposes
on top of the libraries we had created, we had initially thought that one can
use the same approaches as one uses for libraries -- but this turned out to
not work.

Libraries are relatively easy to test because they export large numbers of
functions that one can call individually from small test programs. These are
often called \textit{unit tests} and are typically complemented by integration
tests to verify that combinations of classes work well together. Over the
years we have written about 3,000 of these that are run in multiple
configurations with every single commit to \dealii{} -- on multiple machines,
with different compilers and dependencies, and several times a day. Despite
all of the experience we have gained building this machinery around \dealii{},
the same strategy does not work for whole applications because applications do
not usually export many possible entry points. Rather, they usually only
present a single interface -- the command line, a graphical user interface, or
input files -- and then run through a large fraction of the entire code base
in computing output. One cannot typically just call a single function in an
application by presenting a magic input file, in the same way as one would
call a function in a library by presenting a magic \texttt{main()} function in
a unit test that just sets up some data and then calls the function in
question.

Consequently, different approaches are necessary. In the following, we will
lay out what we have learned from building and testing the \aspect{} code for the
geodynamics community over the past few years. In the following, let us first
talk briefly about what \aspect{} does and how the way it is used is
prototypical for scientific codes (Section~\ref{...}). At the same time,
\aspect{} is a very complex code and just installing it is non-trivial;
consequently, it does not serve the purpose of this book well and we will
present a simple model problem and model code in Section~\ref{...} that we
will then use throughout the rest of the chapter. Section~\ref{...} then
discusses how we can write tests alongside developing the code -- using the
\textit{clone and graft} technique of creating test cases. ...COMPLETE...

\section{\aspect{}: the Advanced Solver for Problems in Earth ConvecTion}

\begin{wrapfigure}{R}{0.44\textwidth}
  \begin{center}
    \vspace*{-24pt}
    \includegraphics[width=0.42\textwidth,height=0.42\textwidth]{figures/aspect.png}
    \vspace*{-12pt}
  \end{center}
  \caption{\it Snapshot of the temperature field of an \aspect{} simulation
    showing convection in the Earth mantle.}
  \vspace*{-3mm}
  \label{fig:aspect}
\end{wrapfigure}
The code that made us think about how to test whole applications is called
\aspect{} -- short for the Advanced Solver for Problems in Earth ConvecTion
\cite{...}. It is a program that simulates how material moves around the Earth
mantle (i.e., the region between the metallic core of Earth and the plates at
the surface on which we live). While the material in the mantle is solid rock,
it is hot and under enormous pressure and can deform with velocities of a few
centimeters per year. On time scales of millions of years, it therefore
behaves like a fluid. Since it is heated from below and cooled from above, it
shows the same kind of behavior as a pot on the stove: blobs of hot material
rise up, cool at the surface, and blobs of cool material fall down. In other
words, it convects. The details of trying to simulate this on a computer go
beyond what we want to discuss here, but it is worth showing a picture and
linking to the videos at \url{http://www.youtube.com/embed/j63MkEc0RRw} and
\url{http://www.youtube.com/embed/EJJ6f4hmDPU} simply because they are pretty.

\aspect{} is a large code: At the time of writing this chapter, it has 297
source files with 69,704 lines of code. What's more, it builds on the
\dealii{} library that has some 500,000 lines of code, the \trilinos{} library
with 3,500,000 lines, and a few other but smaller libraries. At the same time,
this size is not uncommon for large scientific codes, and certainly not for
commercial codes.

Like many academic codes, \aspect{} has no graphical user interface but
instead is driven by input files; it is also designed in such a way that it is
easy to add functionality through plugins, i.e., self-contained source files
that simply implement a class derived from some base class and that is then
registered in \aspect{}'s plugin registry at run time. Because this is the way
users interact with this code, it is also the framework within which we have
to approach testing: tests need to consist of specially crafted input files
and/or plugins.

\section{Make it simple for me: A model problem}

We're not going to try demonstrating our approach using \aspect{} -- that
would be far to complex to install and deal with. Rather, we're going to show
how it works with a much simpler code that just deals with a model problem but
that we will write with the same approach towards input handling and testing
as \aspect{}.

So this is what we're going to consider: Imagine you are dealing with two
baseballs (or, if you're living in the ``rest of the world'', golf balls) that
are connected by a spring. If you throw them, what are their trajectories?

\end{document}

### 2014-11-26 turcksin-spring-mechanics/paper.tex
Write some more.

@@
 are connected by a spring. If you throw them, what are their trajectories?
+Newton's law says that for each of the two balls, mass times acceleration
+equals the force on the body. Let's assume for a moment that there is no air
+friction and that the spring is massless, then the differential equation that
+describes the motion of each of the two bodies ($i=1,2$) is
+\begin{align}
+  \label{eq:ode}
+  m_i 
+  \underbrace{\mathbf x_i''(t)}_{\text{acceleration}}
+  &=
+  \underbrace{m_i \mathbf g}_{\text{gravity}}
+  -
+  \underbrace{D (\|\mathbf x_2(t) - \mathbf x_1(t)\| - L)}_{\text{magnitude of
+    spring force}}
+\underbrace{\frac{\mathbf x_2(t) - \mathbf x_1(t)}{\|\mathbf x_2(t) - \mathbf
+    x_1(t)\|}}_{\text{direction of spring force}},
+\\
+\label{eq:ic1}
+  \mathbf x_i(0) &= \mathbf x_{i,0},
+\\
+\label{eq:ic2}
+  \mathbf x_i'(0) &= \mathbf v_{i,0},
+\end{align}
+where $\mathbf x_i(t)$ is the position of the $i$th body at time $t$, $\mathbf
+g=(0,0,-g)^T$ is the gravity acceleration, $D$ is the spring constant of
+the spring that connects the two balls, and $L$ is the rest length of the
+spring. The second and third equations are the necessary initial conditions
+for this second order differential equation and denote the initial position
+$\mathbf x_{i,0}$ and initial velocity $\mathbf v_{i,0}$.
+
+\section{A first program}
+
+Let's implement a program that can solve these questions. Because we want it
+to be simple and accessible, we're going to use Python as it's widely used and
+because it comes with a number of tools that are going to make our life
+simpler by keeping the program short. Here it is:
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize]
+  from json import *
+  ...
+\end{lstlisting}
+The program reads its input parameters from a file in JSON (see
+\url{http://en.wikipedia.org/wiki/JSON}) format.%
+\footnote{JSON is not the prettiest format to write input files in and we
+  don't recommend you do it for your program (it is typically intended for
+  exchange of data between programs, not between humans and computers). But it
+  will serve for our purposes here primarily because Python has a built-in
+  parser for it that makes our program pleasantly free of the tedium of
+  parsing input files.}
+Here is an example file:
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize]
+  "initial position"   : [ [1,2,3], [4,5,6] ],
+  "initial velocity"   : [ [0,0,0], [1,1,1] ],
+  "masses"             : [ 13.5, 29.75 ],
+  "spring constant"    : 42,
+  "spring rest length" : 2.25
+\end{lstlisting}
+In JSON, the name of a parameter is to the left of the colon and its value is
+to the right; name/value pairs are separated by commas. The value of a
+parameter can be a plain number, or it can be an array if enclosed in
+brackets. Arrays can also be nested, which is what we use here for the initial
+positions and velocities of our two bodies, each of which lives in
+three-dimensional space. With this explanation, it is clear that the input file
+specifies properties of the two masses as $\mathbf x_{1,0}=(1,2,3)^T, \mathbf
+x_{2,0}=(4,5,6)^T, \mathbf v_{1,0}=(0,0,0)^T, \mathbf v_{2,0}=(1,1,1)^T,
+m_1=13.5, m_2=29.75$ and $D=42, L=2.25$ for the spring that connects them, 
+thereby completely determining everything in equations
+\eqref{eq:ode}--\eqref{eq:ic2}.
 

### 2014-11-26 turcksin-spring-mechanics/paper.tex
Write significantly more. Also add some more figures.

@@
     spring force}}
-\underbrace{\frac{\mathbf x_2(t) - \mathbf x_1(t)}{\|\mathbf x_2(t) - \mathbf
-    x_1(t)\|}}_{\text{direction of spring force}},
+\underbrace{\mathbf d_i}_{\text{direction of spring force}},
 \\
@@
 the spring that connects the two balls, and $L$ is the rest length of the
-spring. The second and third equations are the necessary initial conditions
+spring. The direction vector $\mathbf d_i$ is
+$\mathbf d_1 = \frac{\mathbf x_2(t) - \mathbf x_1(t)}{\|\mathbf x_2(t) - \mathbf
+    x_1(t)\|}$
+and
+$\mathbf d_2 = \frac{\mathbf x_1(t) - \mathbf x_2(t)}{\|\mathbf x_1(t) - \mathbf
+    x_2(t)\|}=-\mathbf d_1$ for the two masses.
+The second and third equations are the necessary initial conditions
 for this second order differential equation and denote the initial position
@@
 \begin{lstlisting}[frame=single,basicstyle=\footnotesize]
-  from json import *
+  import json
   ...
@@
 \footnote{JSON is not the prettiest format to write input files in and we
-  don't recommend you do it for your program (it is typically intended for
+  don't recommend you use it for your programs (it is typically intended for
   exchange of data between programs, not between humans and computers). But it
@@
   parsing input files.}
-Here is an example file:
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize]
-  "initial position"   : [ [1,2,3], [4,5,6] ],
-  "initial velocity"   : [ [0,0,0], [1,1,1] ],
-  "masses"             : [ 13.5, 29.75 ],
-  "spring constant"    : 42,
-  "spring rest length" : 2.25
-\end{lstlisting}
+Here is an example file that you can find in the file \texttt{testcase-1.json}:
+\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-1.json}
 In JSON, the name of a parameter is to the left of the colon and its value is
@@
 
+When you run the program on this data, it will compute the trajectory of the
+two connected balls over a number of seconds.
+
+...maybe some output...
+
+\section{Testing for perpetuity}
+
+So how does one test a code like this? There are really two answers to
+this. The first has to do with ensuring that the functionality we have just
+implemented is correct. The second is how we can ensure that the functionality
+we have implemented today will still work correctly tomorrow after
+implementing developing more features.
+
+We often have an intuitive idea of how to do the former. For example, we could
+compare with another code, or we could look at special cases for which we have
+an analytic solution. For our model problem, one could for example prescribe
+initial velocities so that the bodies are at rest and initial positions so
+that the two bodies are separate by the rest length of the spring; the two
+bodies then simply fall straight down and we can easily compute on a piece of
+paper where they should be at time $t$. The following input file
+(\texttt{testcase-2.json}) does this:
+\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-2.json}
+
+In the end, verifying whether the program is correct is problem dependent. Let
+us not go too deep into this because the second answer -- how to ensure that
+functionality continues to work -- is actually much more important in practice
+even though few people realize this when they start developing software. The
+issue essentially boils down to this: I need mechanisms that ensure that
+functionality if I have verified today works correctly, does not break when
+implementing new functionality. This is necessary because good programmers
+understand that they will make mistakes; what makes them good programmers is
+that they develop ways to minimize the impact by helping them find problems
+quickly.
+
+There is a lot of research on bugs in programs. Fundamentally, the essence of
+this research is that it cannot be avoided altogether when programming. On the
+other hand, the best programs are those that make it easy to spot bugs and fix
+them quickly. The worst programs are those that are not tested; an empirically
+true statement is that programs that are not tested produce wrong answers --
+where the emphasis is on the fact that they \textit{produce} wrong answers,
+not that they \textit{may produce} wrong answers. Experienced programmers all
+attest to this statement. Only bad or inexperienced programmers will assume
+that they do not introduce new bugs in old parts of their programs.%
+\footnote{In fact, what seems to set good programmers apart more than many
+  other factor is not that they create fewer bugs -- which they do, though not
+  dramatically so -- but that they understand that they create bugs and that
+  they change their behavior to accommodate for it, for example by developing
+  a habit of rigorous testing that helps them find these bugs quickly.}
+
+The importance of testing relies in the cost of fixing problems. If we keep
+implementing new functionality, eventually some old feature will stop
+working. If we have no tests that detect this quickly, this may go unnoticed
+for a while but eventually we will try to re-run an old computation with the
+new version of the code and it will not work any more. So where to start? What
+have we changed in the intervening months that may cause this? There is likely
+a lot of new code we have written since then, and we may have forgotten about
+a fair share of the details since then. It will take a long time to figure out
+where exactly the problem is. On the other hand, if we had a testsuite that we
+run after essentially every change, we will be notified that something stopped
+working immediately -- with a relatively small number of changes made since
+the test last worked, and everything still fresh on our mind, finding the bug
+should be a much quicker issue.
+
+So how do we achieve this? The general idea is that we save the test cases we
+have previously used to convince ourselves that a new part of the program was
+correct. To make this work, we have to have three things:
+\begin{enumerate}
+\item The input file for a previously verified testcase.
+\item The output when running the code on this input file; having verified the
+  feature, we know or at least believe that this output is correct.
+\item A way to run a new version of the code on the output file, to compare
+  its current output against the stored output, and to raise a flag if they
+  are not the same.
+\end{enumerate}
+The first two are really not very complicated -- create a directory
+\texttt{tests/} in your project and create two files in it, say
+\texttt{testcase-1.json} and \texttt{testcase-1.output}. The last part can
+be as easy as running commands such as
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=csh]
+  python masses-and-springs.py testcase-1.json > testcase-1.output.new
+  diff testcase-1.output testcase-1.output.new
+\end{lstlisting}
+You would repeat all of this for \texttt{testcase-2.json}, of course.
+If the \texttt{diff} program produces no output, then the two files are the
+same and your program computed the same answer on this input file as it did
+back when you created the \texttt{testcase-1.output} file. If there is a
+difference, then you know that you just broke one of the testcases. If you
+purposefully changed it -- for example because you changed the output format,
+or because you fixed behavior you know was erroneous and have verified that it
+is now correct, then copy the new output over the new one. If you did not
+purposefully change it, then you know it's time to break out the debugger.
+
+\section{More functionality}
+
+We write tests because we expect that our code base will continue to change --
+to implement new functionality, or to fix existing bugs. So let's do this:
+assume we now also want to include air friction on the two balls, and we will
+use that the air friction force is proportional to the square of the speed and
+in the opposite direction of the velocity \cite{...}. Then our model
+\eqref{eq:ode} needs to change to the following:
+\begin{align}
+  \label{eq:ode2}
+  m_i 
+  \mathbf x_i''(t)
+  &=
+  m_i \mathbf g
+  -
+  D \left(\|\mathbf x_2(t) - \mathbf x_1(t)\| - L\right) \mathbf d_i
+  \underbrace{-C_i \; \|\mathbf x_i(t)\| \; \mathbf x_i(t)}_{\text{friction force}}.
+\end{align}
+The new term contains constants $C_i$ that denote air friction coefficients of
+the two balls. These coefficients can be computed from the cross section of
+the ball, their surface roughness, the viscosity of the air, and other
+factors, but this is not important here -- we will simply read them from the
+input file.
+
+If we now require that every input file contains an array of friction
+coefficients $C_i$, we can no longer use old input files. A better strategy is
+to define default values for all non-required parameters and overwrite them
+with the values in the input file if specified. A backward compatible choice
+of input parameter for the $C_i$ is to give them a default value of zero
+because in that case equation \eqref{eq:ode2} equals the previous version,
+\eqref{eq:ode}.
+
+Implementing all of this leads to the following program:
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize]
+  ...
+\end{lstlisting}
+
+How do we test this? First, we should check whether the old input files still
+work:
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=csh]
+  python masses-and-springs.py testcase-1.json > testcase-1.output.new
+  diff testcase-1.output testcase-1.output.new
+  python masses-and-springs.py testcase-2.json > testcase-2.output.new
+  diff testcase-2.output testcase-2.output.new
+\end{lstlisting}
+With the code that accompanies this paper, \texttt{diff} reports no difference
+between the old and the new output files, so this is reassuring. Apparently
+the trick with
+
+Next, we need to write new tests for the new functionality. This is where the
+\textit{clone and graft} technique finally comes into play: take one of the
+existing tests (i.e., \textit{clone} it) and add to it the minimal number of
+parameters to exercise the new functionality (i.e., \textit{graft} onto
+it). So let's take the simpler of the two previous testcases,
+\texttt{tests/testcase-2.json}, copy it to \texttt{tests/testcase-3.json} and
+add to it something about air friction: 
+\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-3.json}
+To verify that the results we get are correct, we can use that in this test
+case the masses and friction coefficients of the two balls are the same, the
+two balls are separated by a distance equal to the rest length of the spring,
+and they start at rest. Then they will simply fall as if there was no spring,
+and we know that they will approach a terminal speed $v=\sqrt{mg/C}$ where
+downward gravity and upward air friction cancel. For the current case, this
+would be $v=\sqrt{9.81}$ because we have chosen $m=1, g=9.81, C=1$.
+
+Is one new testcase enough? The answer is rarely yes. For example, it is easy
+to make indexing errors and choosing masses and friction coefficients equal
+would not allow us to detect errors where, for example, we used
+\texttt{friction[1]} in the code where we intended to use \texttt{friction[i]}.
+So let's copy \texttt{tests/testcase-3.json} to
+\texttt{tests/testcase-4.json}:
+\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-4.json}
+Here, the second ball has a greater mass, but the same air friction
+coefficient (think of it as denser but with the same diameter). How do we know
+whether our code is correct for this case as well? This may be more difficult
+to establish, but it isn't always necessary. For example, we could look at the
+output and realize that the second ball falls faster and, because it is
+connected to the first by a spring, then starts to drag the first behind it as
+it falls. Such qualitative verification is sometimes the best one can do. At
+the same time, while it may not guarantee that the code is indeed correct, it
+will nonetheless serve as a useful test in the future to ensure that existing
+functionality does not break as a result of further development -- in the long
+term the more important of the two reasons to write tests.
+
+Because it is so easy to create tests, it may be worthwhile cloning
+\texttt{tests/testcase-3.json} a second time into
+\texttt{tests/testcase-5.json} where now we just set the friction coefficients
+differently:
+\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-4.json}
+Now the two balls have the same mass but the second one has a higher drag
+coefficient (think of the first ball as smooth and the second one as covered
+in fur). Our ``smell check'' would now be to verify that the second ball drags
+behind the first.
+
+\section{Practical considerations}
+
+\subsection{When clone and graft fails}
+
+\begin{wrapfigure}{R}{0.2\textwidth}
+  \begin{center}
+    \vspace*{-24pt}
+    \includegraphics[width=0.1\textwidth]{figures/inheritance.png}
+    \vspace*{-12pt}
+  \end{center}
+  \caption{\it Inheritance graph for the 5 tests discussed in previous sections.}
+  \vspace*{-3mm}
+  \label{fig:inheritance}
+\end{wrapfigure}
+The scheme outlined above represents an inheritance relationship where every
+test is based on a previous one. For the 5 testcases discussed above, the
+inheritance graph is shown in Fig.~\ref{fig:inheritance}. It is, of course, a
+\textit{tree graph}.
+
+But this does not always work. There are cases where you will have to write a
+new test from scratch. To give an example of where this was necessary,
+consider \aspect{}: imagine that at the beginning, it could only compute
+convection in a box. We would have testcases that described different
+temperature boundary conditions on this box, and various other things. But
+when implementing a spherical geometry, none of these testcases quite fit:
+they all had to say what the temperature is on the six faces of the box, they
+made the implicit assumption that gravity points downward instead of inward,
+etc. It was time to start with a new testcase. If, at a later time you then
+clone and graft from this new testcase, then it becomes the root of a separate
+tree (let's call it an \textit{exemplar} testcase) and the inheritance graph
+will turn into a \textit{forest graph} (a collection of tree graphs).
+
+\begin{wrapfigure}{L}{0.28\textwidth}
+  \begin{center}
+    \vspace*{-24pt}
+    \includegraphics[width=0.25\textwidth]{figures/graph.png}
+    \vspace*{-12pt}
+  \end{center}
+  \caption{\it Connection graph of the 144 tests that are part of the
+    \aspect{} code base.}
+  \vspace*{-3mm}
+  \label{fig:aspect-tests}
+\end{wrapfigure}
+Finally, there are cases where you copy-paste pieces from multiple existing
+tests into a new one, for example because you want to test the combination of
+features for which you have already written individual testcases. At this
+point, the metaphor of inheritance graphs starts to break down. However, in
+practice, most tests are still closely related to some others. A visualization
+of this for the 144 tests that are part of \aspect{}, each represented by a
+dot, is shown in Fig.~\ref{fig:aspect-tests} where we draw tests farther apart
+of there are more textual differences between the input files. Many tests come
+in groups of 2, 3 or 4 that only differ in a few lines of input -- simple
+modifications such as those between testcases 3, 4 and 5 above. On the other
+hand, other group are further apart and are derived from different
+examplars.
+
+\subsection{Dealing with many tests}
+
+In practice, once you have accumulated a number of testcases, running and
+comparing them by hand using the commands above becomes unwieldy. In such
+cases, it is useful to write little \texttt{Makefile}s that do this
+automatically, or to use one of the tools available as open source -- for
+example, the \texttt{CTest} program. It also
+becomes more difficult to display tests that create difference when, for
+example, you run the 7,000 tests of the \dealii{} testsuite. Again, there are
+freely available systems such as \texttt{CDash} that present graphical
+overviews of all succeeding and failing tests in a browser.
+
+talk about a better naming scheme to make it easier to find test cases to clone
+
+\subsection{Dealing with round-off}
+
+Talk about numdiff
+
 \end{document}

### 2014-11-26 turcksin-spring-mechanics/figures/graph.sh

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+# to be run in the aspect/tests directory
+
+(echo "graph G {" ; 
+ for i in *prm ; do
+   echo "node [style=filled,color=blue,label=\"\"] `echo $i | sed s/[-_]//g`;" ; 
+ done ; 
+ for i in *prm ; do
+   for j in *prm ; do
+     if test "$i" != "$j" ; then 
+       echo "`echo $i | sed s/[-_]//g` -- `echo $j | sed s/[-_]//g` [style=invis,len=`(echo scale=8 ; echo \`diff $i $j | wc -l\`/10) | bc`];" ; 
+     fi ; 
+   done ;
+ done ; 
+ echo "}"
+) | perl -pi -e 's/\.prm//g;' > graph.dot
+
+neato -Tsvg graph.dot -Gmaxiter=100 -o graph.svg
+neato -Tpng graph.dot -Gmaxiter=100 -o graph.png

### 2014-11-26 turcksin-spring-mechanics/paper.tex
Update graph generation and figures. A bit more text.

@@
 
-\begin{wrapfigure}{L}{0.28\textwidth}
+\begin{wrapfigure}{L}{0.45\textwidth}
   \begin{center}
     \vspace*{-24pt}
-    \includegraphics[width=0.25\textwidth]{figures/graph.png}
+    \includegraphics[height=0.45\textwidth,angle=-90]{figures/graph.png}
     \vspace*{-12pt}
@@
 
+The message from this section, though, is that in almost all cases writing a
+new testcase is actually really easy: find a good starting point, copy it,
+make minor modifications and additions, and verify the output. Testing
+scientific codes really doesn't have to be hard!
+
 \subsection{Dealing with many tests}
 
-In practice, once you have accumulated a number of testcases, running and
-comparing them by hand using the commands above becomes unwieldy. In such
+In practice, once you have accumulated a number of testcases, you will
+encounter practical difficulties. First, when you try to find the right
+starting point to clone and graft a new testcase, you will not recall if
+\texttt{tests/testcase-3.json} or \texttt{tests/testcase-5.json} is the one
+that's closest to the situation you have in mind (for example, if you want to
+augment your model and you are looking again for a testcase with equal masses
+and friction coefficients).
+
+In our experience, using more descriptive filenames for the input files has
+helped, as is placing comments at the top of each input file describing what
+the test does or, if appropriate, how it differs from a previous
+testcase. Many of our testcases in \aspect{} have names like
+\texttt{tests/melt-fraction-peridotite.prm} and comments at the top of the
+form ``\textit{Like the melt-fraction.prm test, but using the peridotite
+  material parameters.}''
+
+Second, running and
+comparing many tests by hand using the commands above becomes unwieldy. In such
 cases, it is useful to write little \texttt{Makefile}s that do this
@@
 
-talk about a better naming scheme to make it easier to find test cases to clone
-
 \subsection{Dealing with round-off}
@@
 
+\section{Conclusions}
+
 \end{document}

### 2014-11-26 turcksin-spring-mechanics/figures/graph.sh

@@ -7,9 +7,11 @@
    echo "node [style=filled,color=blue,label=\"\"] `echo $i | sed s/[-_]//g`;" ; 
  done ; 
  for i in *prm ; do
+   cat $i | perl -pi -e 's/^[ \t]*\n//g; s/^\#.*\n//g;' > 1
    for j in *prm ; do
      if test "$i" != "$j" ; then 
-       echo "`echo $i | sed s/[-_]//g` -- `echo $j | sed s/[-_]//g` [style=invis,len=`(echo scale=8 ; echo \`diff $i $j | wc -l\`/10) | bc`];" ; 
+       cat $j | perl -pi -e 's/^[ \t]*\n//g; s/^\#.*\n//g;' > 2
+       echo "`echo $i | sed s/[-_]//g` -- `echo $j | sed s/[-_]//g` [style=invis,len=`(echo scale=8 ; echo \`diff 1 2 | wc -l\`/10) | bc`];" ; 
      fi ; 
    done ;
  done ; 

### 2014-12-02 turcksin-spring-mechanics/paper.tex
Write a few more sentences.

@@
 
-%\runningheads{B. Turcksin, T. Heister, W. Bangerth}{Clone and graft: Testing whole applications}
+%\runningheads{B. Turcksin, T. Heister, W. Bangerth}{Clone and graft: Testing applications}
 
-\title{Clone and graft:\\ Testing whole scientific
+\title{Clone and graft:\\ Testing scientific
   applications as they are built}
@@
 
-Second, running and
-comparing many tests by hand using the commands above becomes unwieldy. In such
-cases, it is useful to write little \texttt{Makefile}s that do this
-automatically, or to use one of the tools available as open source -- for
-example, the \texttt{CTest} program. It also
-becomes more difficult to display tests that create difference when, for
+Second, running and comparing many tests by hand using the commands above
+becomes unwieldy. In such cases, it is useful to write little
+\texttt{Makefile}s that do this automatically, or to use one of the tools
+available as open source -- for example, the \texttt{CTest} program. It also
+becomes more difficult to display tests that create differences when, for
 example, you run the 7,000 tests of the \dealii{} testsuite. Again, there are
@@
 
+In practice, you're not likely to quickly get to a project size where learning
+to use tools like \texttt{CTest} and \texttt{CDash} are important
+considerations. In practice, the point where using such tools becomes
+indispensible is when your code has several 10,000 lines of code or more --
+several years worth of work. When you get to this point, your project will
+have all sorts of other growing pains as well, such as having to deal with
+external users, running mailing lists, portability, managing complexity and
+other issues. Some of these are discussed in \cite{BH13}. The point simply is
+that there is no need to worry too much about testing tools for programs of
+this size: care about writing tests from the beginning and if the number of
+tests becomes too unwieldy, there will be professional-grade tools you can
+switch to that can manage this for you. In the meantime, a little shell script
+such as the following will do just fine:
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=bash]
+  # loop over all .json input files
+  for jsonfile in *.json ; do
+    # execute our program on it and pipe the output into a file
+    python ${jsonfile} > ${jsonfile}.new
+
+    # then compare output; only ask 'diff' to report whether the
+    # files are the same or not, and create output depending on this:
+    # print an 'X' if there is a difference, a '.' if there isn't. this
+    # ensures that test failures are easily visible
+    if (diff ${jasonfile} ${jsonfile}.new > /dev/null) ; then
+      echo " .    ${jasonfile}" ;
+    else 
+      echo " X    ${jasonfile}" ;
+    fi
+  done
+\end{lstlisting}
+If you call this script, say, \texttt{tests/compare.sh}, then running it using
+the command \texttt{cd tests/ ; bash -e compare.sh} should produce output like
+this:
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=bash]
+ .    testcase-1.json
+ .    testcase-2.json
+ .    testcase-3.json
+ .    testcase-4.json
+ .    testcase-5.json
+\end{lstlisting}
+On the other hand, if your output looks like this
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=bash]
+ .    testcase-1.json
+ .    testcase-2.json
+ .    testcase-3.json
+ X    testcase-4.json
+ .    testcase-5.json
+\end{lstlisting}
+then you know that you just broke some functionality that
+\texttt{testcase-4.json} checks. Given how easy it is to run this script, you
+want to run this script often to find new bugs as early as possible.
+
 \subsection{Dealing with round-off}
+
+Scientific computing deals, at a fundamental level, almost always with
+floating point numbers. Unfortunately, floating point arithmetic suffers from
+some rather annoying oddities that can make testing scientific codes more
+challenging than code that doesn't use floating point numbers.
+
+First, floating point numbers do not enjoy the same properties as integers.
 

### 2014-12-03 turcksin-spring-mechanics/paper.tex
More.

@@
 
-First, floating point numbers do not enjoy the same properties as integers.
+At the base of the problem is the fact that floating point numbers do not
+enjoy the same mathematical properties as integers (or, in fact, as real
+numbers). For example, if you take into account round-off, then $(a+b)+c$ is in
+general \textit{not} the same as $a+(b+c)$ or $(a+c)+b$. You can try this out
+with $a=1.0, b=-1.0, c=1e-20$: in double precision, you will get
+$(a+b)+c=1e-20, a+(b+c)=(a+c)+b=0$. This matters because if you make changes
+to the code base, upgrade the compiler, switch from one processor or machine
+to another, or change compiler optimization level, floating point operations
+will be executed in different order and the results will be different at the
+level of roundoff. In other words, your tests will start to fail even though
+nothing fundamentally changed -- but the output will be different.
 
-Talk about numdiff
+Over the past 15 years, we have played with many ideas how to address this
+problem. For example, we tried to just never output numbers to their full 16
+digits of double precision. The idea is that if you only output 5 digits, then
+roundoff surely can't play a role. The problem is that the numbers 1.000\,049
+999\,999\,999\,9 and 1.000\,050\,000\,000\,000\,0 only differ by round-off,
+but when rounded to 5 digits, you either get 1.000\,0 or 1.000\,1. One could
+think that this is a contrived example, but it turns out that this happens
+with rather frustrating regularity if you just output enough numbers -- as one
+often does in scientific codes.
+
+A second area that creates problems that codes often try to compute zeros, but
+because of roundoff what is actually produced are numbers of the kind
+4.863e-19. A human recognizes that this is a zero and, consequently, equal to
+2.489e-19 if all of the numbers around it are on the order of one. But the
+\texttt{diff} command will complain, and consequently signal a difference.
+
+The solution to these problems that we have ultimately come to use
+is a program called \texttt{numdiff}. \texttt{numdiff} works essentially like
+\texttt{diff} but it recognizes if the text it compares in two files is
+actually a number. If two numbers differ by less than some relative or
+absolute tolerance, then they are considered equal even if the characters by
+which they are represented in an output file are different. For example, you
+would replace the \texttt{diff} command used previously by the
+following to ensure that numbers $a$ and $b$ compare as equal if $|a-b|\le
+10^{-6}$ or $|a-b|\le 10^{-8} \max\{|a|,|b|\}$:
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=csh]
+  numdiff -a 1e-6 -r 1e-8 testcase-1.output testcase-1.output.new
+\end{lstlisting}
 

### 2014-12-03 turcksin-spring-mechanics/paper.tex
Provide cross-references within the paper.

@@
 talk briefly about what \aspect{} does and how the way it is used is
-prototypical for scientific codes (Section~\ref{...}). At the same time,
+prototypical for scientific codes (Section~\ref{sec:aspect}). At the same time,
 \aspect{} is a very complex code and just installing it is non-trivial;
 consequently, it does not serve the purpose of this book well and we will
-present a simple model problem and model code in Section~\ref{...} that we
-will then use throughout the rest of the chapter. Section~\ref{...} then
-discusses how we can write tests alongside developing the code -- using the
-\textit{clone and graft} technique of creating test cases. ...COMPLETE...
+present a simple model problem and model code in
+Sections~\ref{sec:model-problem} and \ref{sec:first-steps} that we
+will then use throughout the rest of the chapter. Sections~\ref{sec:testing}
+and \ref{sec:extending} then
+discuss how we write tests for whole applications, and how this process
+happens as we continue to develop the code -- using the
+\textit{clone and graft} technique of creating test
+cases. Section~\ref{sec:practice} then deals with how real practice interferes
+with the ideas discussed in previous sections, and we conclude and summarize
+in Section~\ref{sec:conclusions}.
 
 \section{\aspect{}: the Advanced Solver for Problems in Earth ConvecTion}
+\label{sec:aspect}
 
@@
 to approach testing: tests need to consist of specially crafted input files
-and/or plugins.
+and/or plugins. Plugins go beyond what we can discuss here, but we will show
+you that it is actually quite simple to write tests using input files.
 
 \section{Make it simple for me: A model problem}
+\label{sec:model-problem}
 
@@
 \section{A first program}
+\label{sec:first-steps}
 
@@
 \section{Testing for perpetuity}
+\label{sec:testing}
 
@@
 \section{More functionality}
+\label{sec:extending}
 
@@
 \section{Practical considerations}
+\label{sec:practice}
 
@@
 \section{Conclusions}
+\label{sec:conclusions}
 

### 2014-12-03 turcksin-spring-mechanics/paper.tex
Finish writing the text, add bibliography.

@@
 library for finite element computations (see
-\url{http://www.dealii.org/}). Libraries are of course the foundation of
+\url{http://www.dealii.org/} and \cite{BHK07,BK99m}). Libraries are of course the foundation of
 almost every single scientific code, starting with ``simple'' ones such as
 BLAS and LAPACK to very much more complex ones such as \petsc{}, \trilinos{}, or
-\dealii{}.
+\dealii{} \cite{petsc,trilinos,trilinos-web-page}.
 
@@
 \aspect{} -- short for the Advanced Solver for Problems in Earth ConvecTion
-\cite{...}. It is a program that simulates how material moves around the Earth
+\cite{KHB12}. It is a program that simulates how material moves around the Earth
 mantle (i.e., the region between the metallic core of Earth and the plates at
@@
 digits of double precision. The idea is that if you only output 5 digits, then
-roundoff surely can't play a role. The problem is that the numbers 1.000\,049
-999\,999\,999\,9 and 1.000\,050\,000\,000\,000\,0 only differ by round-off,
-but when rounded to 5 digits, you either get 1.000\,0 or 1.000\,1. One could
+roundoff surely can't play a role. The problem is that the numbers 1.234\,549
+999\,999\,999\,9 and 1.234\,550\,000\,000\,000\,1 only differ by round-off,
+but when rounded to 5 digits, you either get 1.234\,5 or 1.234\,6. One could
 think that this is a contrived example, but it turns out that this happens
@@
 following to ensure that numbers $a$ and $b$ compare as equal if $|a-b|\le
-10^{-6}$ or $|a-b|\le 10^{-8} \max\{|a|,|b|\}$:
+10^{-6}$ or $|a-b|\le 10^{-8} \min\{|a|,|b|\}$:
 \begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=csh]
@@
 \end{lstlisting}
+Using this trick, numbers that are written into the output file
+\texttt{testcase-1.output.new} on one machine with slight differences to the
+ones originally saved in \texttt{testcase-1.output} will still be considered
+the same, and the comparison will succeed.
 
-\section{Conclusions}
+\section{Summary}
 \label{sec:conclusions}
 
+When we talk to students who are just starting writing their own codes, then
+they typically understand the importance of testing in some abstract
+sense. But almost all of them have two fundamental misconceptions: First, that
+their codes are either written well enough or will not be changed in ways
+fundamental enough to suggest that these changes might break existing
+functionality. And second, that doing adequate testing is just too hard and
+cumbersome to be a justifiable expense of time.
+
+The first of these two points is empirically false. Only bad programmers
+expect that they will not break existing functionality, no matter how trivial
+a change might be. All experienced programmers will know this to be wrong. The
+reason is that over the course of a scientific code -- say, over the course of
+the 3 years a graduate student does research, and maybe even beyond this if
+the code is handed on to the next student, or if you go on to be a postdoc --
+a code starts out small but is constantly developed into something much larger
+with much more functionality in all sorts of directions. You're not going to
+create one essential version of the code that you can verify to be correct and
+then just round off some edges; you're in fact going to replace the linear by
+a nonlinear model, replace the direct solver by an iterative one, add terms to
+the equation, and make other very fundamental changes that will require you to
+edit all throughout your code. There is no other way than systematic testing
+to ensure that the computation you ran today will still run tomorrow. And you
+will have to run it again tomorrow: your adviser or boss may ask you to redo a
+computation with some slight changes, a reviewer of a paper might ask to do
+additional simulations, you may have forgotten the exact parameters for a data
+set you want to use in your thesis and will have to redo it to verify,
+etc. You may think that once you have created the data 
+for some particular simulation, that the issue is done and will not have to
+repeated -- i.e., that you might in fact just throw away the code -- but this
+is not typically true.
+
+The second of the poinst above -- that testing is hard and time consuming --
+is also empirically false. Once you start to do it systematically, all you
+have to do is morph an existing input file into a new one that tests the
+aspect you have just implemented, and add it to your testsuite. It doesn't
+take very long, you just have to do it right from the start. Once you do this,
+you will find how much faster it suddenly becomes to debug codes. To us, one
+of the most surprising facts of developing stringent testing procedures is not
+just how much better our codes become, but in fact how much safer we feel
+making changes in central places of our code: without tests, we always felt
+insecure and wondered what a change might do to the code, whether it is
+correct in all cases, and if we might be forgetting some cases. With tests,
+we find ourselves a lot more at ease: just make a change, run the 140 or so
+tests and see what happens -- if it all continues to work, then that may not
+be a guarantee that a change is correct, but it sure makes you sleep better at
+night. It really isn't all that hard to find peace of mind!
+
+\bibliographystyle{alpha}
+\bibliography{paper}
+
 \end{document}

### 2014-12-07 turcksin-spring-mechanics/spring.py
simple ODE solver example

@@ -5,3 +5,37 @@ with open("settings.json", 'r') as f:
 
 
 print settings["spring_constant"]
+
+
+
+# example for ODE solver:
+
+from scipy.integrate import ode
+import numpy as np
+import matplotlib.pyplot as plt
+
+y0 = [80, 30]
+t0 = 0
+T = 100
+
+def f(t, y):
+    return [0.125*y[0]-0.01*y[0]*y[1], -y[1]+0.01*y[0]*y[1]]
+
+r = ode(f).set_integrator('vode',rtol=1e-8)
+r.set_initial_value(y0, t0)
+
+tarr = []
+y1arr = []
+y2arr = []
+
+while r.successful() and r.t < T:
+    r.integrate(T, step=True)
+    tarr.append(r.t)
+    y1arr.append(r.y[0])
+    y2arr.append(r.y[1])
+
+print len(tarr)
+
+#plt.plot(tarr, y1arr, 'yo-', tarr, y2arr, 'rx-')
+plt.plot(y1arr, y2arr, 'r-')
+plt.show()

### 2014-12-07 turcksin-spring-mechanics/spring.py
solve the correct problem (hard-coded constants)

@@ -14,28 +14,52 @@ from scipy.integrate import ode
 import numpy as np
 import matplotlib.pyplot as plt
 
-y0 = [80, 30]
-t0 = 0
-T = 100
+D = 52.
+L = 2.25
+g = [0.,0.,-9.81]
+
+# variable order:
+# p1x p1y p1z p2x p2y p2z v1x v1y v1z v2x v2y v2z
+
+y0 = [1.,2.,3., 4.,5.,6., 0.,0.,0., 1.,1.,1.]
+t0 = 0.
+T = 0.5
+
+m = [3.0, 1.0]
+
 
 def f(t, y):
-    return [0.125*y[0]-0.01*y[0]*y[1], -y[1]+0.01*y[0]*y[1]]
+    p1 = y[0:3]
+    p2 = y[3:6]
+    v1 = y[6:9]
+    v2 = y[9:12]
+    
+    dist = np.linalg.norm(p2-p1)
+    a1 = g - D*(dist-L) * (p2-p1)/dist/m[0];
+    a2 = g - D*(dist-L) * (p1-p2)/dist/m[1];
+    return np.concatenate([v1, v2, a1, a2])             
 
-r = ode(f).set_integrator('vode',rtol=1e-8)
+r = ode(f).set_integrator('vode',rtol=1e-6)
 r.set_initial_value(y0, t0)
 
-tarr = []
-y1arr = []
-y2arr = []
+t = [t0]
+yarr = np.array([y0])
 
 while r.successful() and r.t < T:
     r.integrate(T, step=True)
-    tarr.append(r.t)
-    y1arr.append(r.y[0])
-    y2arr.append(r.y[1])
+    t.append(r.t)
+    yarr = np.vstack((yarr,r.y))
 
-print len(tarr)
+print "time steps:",len(t)
 
-#plt.plot(tarr, y1arr, 'yo-', tarr, y2arr, 'rx-')
-plt.plot(y1arr, y2arr, 'r-')
+
+plt.plot(t, yarr[:,0],
+         t, yarr[:,1],
+         t, yarr[:,2],
+         t, yarr[:,3],
+         t, yarr[:,4],
+         t, yarr[:,5])
+
+plt.legend(["p1.x","p1.y", "..."])
+
 plt.show()

### 2014-12-08 turcksin-spring-mechanics/spring.py
Make the plot 3d. Fix a sign error.

@@ -13,6 +13,7 @@ print settings["spring_constant"]
 from scipy.integrate import ode
 import numpy as np
 import matplotlib.pyplot as plt
+from mpl_toolkits.mplot3d import Axes3D
 
 D = 52.
 L = 2.25
@@ -23,7 +24,7 @@ g = [0.,0.,-9.81]
 
 y0 = [1.,2.,3., 4.,5.,6., 0.,0.,0., 1.,1.,1.]
 t0 = 0.
-T = 0.5
+T = 2
 
 m = [3.0, 1.0]
 
@@ -35,8 +36,8 @@ def f(t, y):
     v2 = y[9:12]
     
     dist = np.linalg.norm(p2-p1)
-    a1 = g - D*(dist-L) * (p2-p1)/dist/m[0];
-    a2 = g - D*(dist-L) * (p1-p2)/dist/m[1];
+    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0];
+    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1];
     return np.concatenate([v1, v2, a1, a2])             
 
 r = ode(f).set_integrator('vode',rtol=1e-6)
@@ -53,13 +54,9 @@ while r.successful() and r.t < T:
 print "time steps:",len(t)
 
 
-plt.plot(t, yarr[:,0],
-         t, yarr[:,1],
-         t, yarr[:,2],
-         t, yarr[:,3],
-         t, yarr[:,4],
-         t, yarr[:,5])
-
-plt.legend(["p1.x","p1.y", "..."])
+fig = plt.figure()
+ax = fig.gca(projection='3d')
+ax.plot(yarr[:,0], yarr[:,1], yarr[:,2], label='body 1')
+ax.plot(yarr[:,3], yarr[:,4], yarr[:,5], label='body 2')
 
 plt.show()

### 2014-12-08 turcksin-spring-mechanics/paper.tex
Add a footnote about the sub-packages that need to be installed.

@@
 because it comes with a number of tools that are going to make our life
-simpler by keeping the program short. Here it is:
+simpler by keeping the program short.%
+\footnote{In order to run this program, you will have to have the Python
+  NumPy, SciPy, and MatPlotLib packages installed on your system, along with
+  the basic Python interpreter.}
+Here it is:
 \begin{lstlisting}[frame=single,basicstyle=\footnotesize]
@@
   parsing input files.}
-Here is an example file that you can find in the file \texttt{testcase-1.json}:
+Here is an example that you can find in the file \texttt{testcase-1.json}:
 \lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-1.json}

### 2014-12-08 turcksin-spring-mechanics/spring.py
Parse json input file.

@@ -1,11 +1,10 @@
+# read input file
+import sys
 import json
 
-with open("settings.json", 'r') as f:
+assert len(sys.argv)==2, 'Please provide an input file.'
+with open(sys.argv[1], 'r') as f:
     settings = json.loads(f.read())
-
-
-print settings["spring_constant"]
-
 
 
 # example for ODE solver:
@@ -15,18 +14,23 @@ import numpy as np
 import matplotlib.pyplot as plt
 from mpl_toolkits.mplot3d import Axes3D
 
-D = 52.
-L = 2.25
-g = [0.,0.,-9.81]
+D = settings['spring constant']
+L = settings['spring rest length']
+g = [0., 0., -9.81]
 
 # variable order:
 # p1x p1y p1z p2x p2y p2z v1x v1y v1z v2x v2y v2z
 
-y0 = [1.,2.,3., 4.,5.,6., 0.,0.,0., 1.,1.,1.]
+init_p1 = settings['initial position'][0]
+init_p2 = settings['initial position'][1]
+init_v1 = settings['initial velocity'][0]
+init_v2 = settings['initial velocity'][1]
+y0 = [init_p1[0], init_p1[1], init_p1[2], init_p2[0], init_p2[1], init_p2[2],\
+        init_v1[0], init_v1[1], init_v1[2], init_v2[0], init_v2[1], init_v2[2]]
 t0 = 0.
 T = 2
 
-m = [3.0, 1.0]
+m = [settings['masses'][0], settings['masses'][1]]
 
 
 def f(t, y):
@@ -34,13 +38,13 @@ def f(t, y):
     p2 = y[3:6]
     v1 = y[6:9]
     v2 = y[9:12]
-    
-    dist = np.linalg.norm(p2-p1)
-    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0];
-    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1];
-    return np.concatenate([v1, v2, a1, a2])             
 
-r = ode(f).set_integrator('vode',rtol=1e-6)
+    dist = np.linalg.norm(p2-p1)
+    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0]
+    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1]
+    return np.concatenate([v1, v2, a1, a2])
+
+r = ode(f).set_integrator('vode', rtol=1e-6)
 r.set_initial_value(y0, t0)
 
[…]

### 2014-12-10 turcksin-spring-mechanics/paper.tex
Change a couple of sentences.

@@
 So this is what we're going to consider: Imagine you are dealing with two
-baseballs (or, if you're living in the ``rest of the world'', golf balls) that
+baseballs that
 are connected by a spring. If you throw them, what are their trajectories?
@@
 we have implemented today will still work correctly tomorrow after
-implementing developing more features.
+implementing more features.
 
@@
 With the code that accompanies this paper, \texttt{diff} reports no difference
-between the old and the new output files, so this is reassuring. Apparently
-the trick with
+between the old and the new output files, so this is reassuring.
 

### 2014-12-10 turcksin-spring-mechanics/spring.py
Add friction to the code.

@@ -18,6 +18,15 @@ D = settings['spring constant']
 L = settings['spring rest length']
 g = [0., 0., -9.81]
 
+# Try to read "air friction coefficient". If it does not exist C1 and C2 are set
+# to zero
+try:
+    C1 = settings['air friction coefficient'][0]
+    C2 = settings['air friction coefficient'][1]
+except KeyError:
+    C1 = 0.
+    C2 = 0.
+
 # variable order:
 # p1x p1y p1z p2x p2y p2z v1x v1y v1z v2x v2y v2z
 
@@ -40,8 +49,8 @@ def f(t, y):
     v2 = y[9:12]
 
     dist = np.linalg.norm(p2-p1)
-    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0]
-    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1]
+    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0] - C1*np.linalg.norm(p1)*p1
+    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1] - C2*np.linalg.norm(p2)*p2
     return np.concatenate([v1, v2, a1, a2])
 
 r = ode(f).set_integrator('vode', rtol=1e-6)

### 2014-12-12 turcksin-spring-mechanics/spring.py
Change indentation to comply to pep8.

@@ -2,7 +2,7 @@
 import sys
 import json
 
-assert len(sys.argv)==2, 'Please provide an input file.'
+assert len(sys.argv) == 2, 'Please provide an input file.'
 with open(sys.argv[1], 'r') as f:
     settings = json.loads(f.read())
 
@@ -18,8 +18,8 @@ D = settings['spring constant']
 L = settings['spring rest length']
 g = [0., 0., -9.81]
 
-# Try to read "air friction coefficient". If it does not exist C1 and C2 are set
-# to zero
+# Try to read "air friction coefficient". If it does not exist C1 and C2 are
+# set to zero
 try:
     C1 = settings['air friction coefficient'][0]
     C2 = settings['air friction coefficient'][1]
@@ -34,8 +34,8 @@ init_p1 = settings['initial position'][0]
 init_p2 = settings['initial position'][1]
 init_v1 = settings['initial velocity'][0]
 init_v2 = settings['initial velocity'][1]
-y0 = [init_p1[0], init_p1[1], init_p1[2], init_p2[0], init_p2[1], init_p2[2],\
-        init_v1[0], init_v1[1], init_v1[2], init_v2[0], init_v2[1], init_v2[2]]
+y0 = [init_p1[0], init_p1[1], init_p1[2], init_p2[0], init_p2[1], init_p2[2],
+      init_v1[0], init_v1[1], init_v1[2], init_v2[0], init_v2[1], init_v2[2]]
 t0 = 0.
 T = 2
 

### 2014-12-19 turcksin-spring-mechanics/paper.tex
Add another section.

@@
 Here it is:
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize]
-  import json
-  ...
-\end{lstlisting}
+\marginpar{After documenting spring.py, split it and simplify the first
+  version to the pieces we need here}
+\lstinputlisting[frame=single,basicstyle=\footnotesize,language=Python]{spring.py}
 The program reads its input parameters from a file in JSON (see
@@
   parsing input files.}
-Here is an example that you can find in the file \texttt{testcase-1.json}:
+Here is an example that you can find in the file \texttt{tests/testcase-1.json}:
 \lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-1.json}
@@
 Implementing all of this leads to the following program:
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize]
-  ...
-\end{lstlisting}
+\lstinputlisting[frame=single,basicstyle=\footnotesize,language=Python]{spring.py}
 
-How do we test this? First, we should check whether the old input files still
+How do we test this new functionality? First, we should check whether the old input files still
 work:
@@
 \subsection{Dealing with many tests}
+\label{sec:many-tests}
 
@@
 want to run this script often to find new bugs as early as possible.
+
+\subsection{Dealing with overlapping tests}
+
+In an ideal world, every testcase would only check a single feature (such
+testcases are then often called \textit{unit tests}). This way, if something
+goes wrong with a change to the source code, a single test will fail and it
+will be easy to identify what went wrong.
+
+In practice, this is often difficult to achieve. For example, if we change the
+implementation of the function that provides $f(t,y)$ above and break it, it
+is likely that \textit{all} testcases will start to fail. Similarly, if you
+build testcases by cloning and grafting onto existing ones, whenever the old
+one fails, the new one will likely do so as well.
+
+From experience, it turns out that this is not often a real problem. The vast
+majority of patches concern just small, isolated pieces of the software and in
+those cases, only one or a few testcases will start failing. It is then easy
+to identify the simplest one of these (assuming that they were given useful
+names, see above in Section~\ref{sec:many-tests}), and that is the one one
+will then use for debugging. If one breaks fundamental functionality somewhere
+in the core of the software, a large number of tests will start to fail
+but in those cases you will either be able to quickly find out what is
+happening, or be able to debug the problem by choosing \textit{any} simple
+test that was among those that broke. In summary, from a practical
+perspective, we have found that it is not usually a problem to have testcases
+that check overlapping parts of the software.
 

### 2015-01-14 turcksin-spring-mechanics/paper.tex
Some more.

@@
 
+\marginpar{Bruno - please update email address}
 \section{Introduction}
@@
 
-Libraries are relatively easy to test because they export large numbers of
-functions that one can call individually from small test programs. These are
+Libraries are relatively easy to test. They export large numbers of
+functions; one can write a simple \texttt{main()} function that sets up a
+couple of data structures, calls one of these functions individually, and
+verifies the correctness of the output; do
+this for every function or class in the library and you've got a good
+testsuite. Such tests are
 often called \textit{unit tests} and are typically complemented by integration
@@
 call a function in a library by presenting a magic \texttt{main()} function in
-a unit test that just sets up some data and then calls the function in
-question.
+a unit test.
 
@@
 will then use throughout the rest of the chapter. Sections~\ref{sec:testing}
-and \ref{sec:extending} then
+and \ref{sec:extending}
 discuss how we write tests for whole applications, and how this process
@@
 \aspect{} -- short for the Advanced Solver for Problems in Earth ConvecTion
-\cite{KHB12}. It is a program that simulates how material moves around the Earth
+\cite{KHB12,aspectmanual,aspectweb}. It is a program that simulates how material moves around the Earth
 mantle (i.e., the region between the metallic core of Earth and the plates at
@@
 linking to the videos at \url{http://www.youtube.com/embed/j63MkEc0RRw} and
-\url{http://www.youtube.com/embed/EJJ6f4hmDPU} simply because they are pretty.
+\url{http://www.youtube.com/embed/EJJ6f4hmDPU}, of only because they are pretty.
 
@@
 
-Let's implement a program that can solve these questions. Because we want it
+Let's implement a program that can solve this problem. Because we want it
 to be simple and accessible, we're going to use Python as it's widely used and
@@
   the basic Python interpreter.}
-Here it is:
-\marginpar{After documenting spring.py, split it and simplify the first
-  version to the pieces we need here}
-\lstinputlisting[frame=single,basicstyle=\footnotesize,language=Python]{spring.py}
-The program reads its input parameters from a file in JSON (see
-\url{http://en.wikipedia.org/wiki/JSON}) format.%
+The program we'll show you in a second reads its input parameters from a file in JSON format%
 \footnote{JSON is not the prettiest format to write input files in and we
@@
   parser for it that makes our program pleasantly free of the tedium of
-  parsing input files.}
+  parsing input files.} (see
+\url{http://en.wikipedia.org/wiki/JSON}).
 Here is an example that you can find in the file \texttt{tests/testcase-1.json}:
-\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-1.json}
+\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-1.json}
 In JSON, the name of a parameter is to the left of the colon and its value is
@@
 \eqref{eq:ode}--\eqref{eq:ic2}.
+
+The program that can deal with this input looks like this:
+\marginpar{After documenting spring.py, split it and simplify the first
+  version to the pieces we need here}
+\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]{spring.py}
 
@@
 be as easy as running commands such as
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=csh]
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
   python masses-and-springs.py testcase-1.json > testcase-1.output.new
@@
 work:
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=csh]
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
   python masses-and-springs.py testcase-1.json > testcase-1.output.new
@@
 such as the following will do just fine:
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=bash]
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]
   # loop over all .json input files
@@
 this:
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=bash]
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]
  .    testcase-1.json
@@
 On the other hand, if your output looks like this
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=bash]
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]
  .    testcase-1.json
@@
 10^{-6}$ or $|a-b|\le 10^{-8} \min\{|a|,|b|\}$:
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize,language=csh]
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
   numdiff -a 1e-6 -r 1e-8 testcase-1.output testcase-1.output.new
@@
 
+\marginpar{Acknowledgments}
+
 \bibliographystyle{alpha}

### 2015-01-14 turcksin-spring-mechanics/paper.tex
Add acknowledgments.

@@
 
-\marginpar{Acknowledgments}
+\paragraph*{Acknowledgments}
+B.~Turcksin and W.~Bangerth were partially supported by the National Science
+Foundation under award OCI-1148116 as part of the Software Infrastructure for
+Sustained Innovation (SI2) program; by the Computational
+Infrastructure in Geodynamics initiative (CIG), through the National Science
+Foundation under Award No.~EAR-0949446 and The University of California --
+Davis.
+\marginpar{Timo -- please update.}
 

### 2015-01-15 turcksin-spring-mechanics/paper.tex
Fix a typo.
Move forward some.

@@
 equals the force on the body. Let's assume for a moment that there is no air
-friction and that the spring is massless, then the differential equation that
-describes the motion of each of the two bodies ($i=1,2$) is
+friction and that the spring is massless, then the ordinary differential
+equation (ODE) that describes the motion of each of the two bodies ($i=1,2$) is
 \begin{align}
@@
 
-The program that can deal with this input looks like this:
+To implement a program that can deal with this input, we first need to
+describe how Python solves ODEs. Basically, it requires us to reformulate it
+as a first order system $\mathbf y'(t)=\mathbf f(t,\mathbf y), \mathbf
+y(0)=\mathbf y_0$, and tell the integrator about $\mathbf f$ and $\mathbf
+y_0$. We do this by introducing variables $\mathbf v_i=\mathbf x'_i$ and then
+setting $\mathbf
+y=(x_{1,1},x_{1,2},\ldots,x_{2,3},v_{1,1},v_{1,2},\ldots,v_{2,3})^T$. $\mathbf
+f$ then consists of the $\mathbf v_i$ and the right hand sides of
+\eqref{eq:ode} divided by the $m_i$. With this, the code looks like this:
 \marginpar{After documenting spring.py, split it and simplify the first
@@
 
+\begin{wrapfigure}{R}{0.2\textwidth}
+  \begin{center}
+    \vspace*{-24pt}
+    \includegraphics[width=0.18\textwidth]{tests/testcase-1.png}
+    \vspace*{-12pt}
+  \end{center}
+  \caption{\it Trajectories for the two balls with the input file from
+    Section~\ref{sec:first-steps}. Note that axes are not shown at the same
+    scale.}
+  \vspace*{-3mm}
+  \label{fig:testcase-1}
+\end{wrapfigure}
+
 When you run the program on this data, it will compute the trajectory of the
-two connected balls over a number of seconds.
-
-...maybe some output...
+two connected balls over a number of seconds and print their final positions
+to the console.
+Of course, having a plot of their trajectories would be nice too, and you can
+get that by adding the following lines to the bottom of the script:
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
+import matplotlib.pyplot
+from mpl_toolkits.mplot3d import Axes3D
+fig = matplotlib.pyplot.figure()
+canvas = fig.gca(projection='3d')
+canvas.plot(y_values[:,0], y_values[:,1], y_values[:,2], label='body 1')
+canvas.plot(y_values[:,3], y_values[:,4], y_values[:,5], label='body 2')
+matplotlib.pyplot.show()
+\end{lstlisting}
+The output this creates is shown in Fig.~\ref{fig:testcase-1}.
 
@@
 We often have an intuitive idea of how to do the former. For example, we could
-compare with another code, or we could look at special cases for which we have
+note that the center of mass at the end time $T=5$ should have fallen a
+distance of $\delta z=\frac 12 g T^2=\frac{9.81 \cdot 25}{2}\textrm m\approx
+123\textrm{m}$, well in agreement with what Fig.~\ref{fig:testcase-1} shows. We
+could also compare with another code. Or we could look at special cases for
+which we have 
 an analytic solution. For our model problem, one could for example prescribe
 initial velocities so that the bodies are at rest and initial positions so
-that the two bodies are separate by the rest length of the spring; the two
+that the two bodies are separated by the rest length of the spring; the two
 bodies then simply fall straight down and we can easily compute on a piece of
@@
 
-The second of the poinst above -- that testing is hard and time consuming --
+The second of the points above -- that testing is hard and time consuming --
 is also empirically false. Once you start to do it systematically, all you

### 2015-01-15 turcksin-spring-mechanics/spring.py

@@ -1,46 +1,37 @@
-# read input file
 import sys
 import json
+import numpy
 
+# Read the name of the input file from the command line, and read options from the file:
 assert len(sys.argv) == 2, 'Please provide an input file.'
 with open(sys.argv[1], 'r') as f:
     settings = json.loads(f.read())
 
-
-# example for ODE solver:
-
-from scipy.integrate import ode
-import numpy as np
-import matplotlib.pyplot as plt
-from mpl_toolkits.mplot3d import Axes3D
-
+# Then retrieve the various parameters from what we just read:
 D = settings['spring constant']
 L = settings['spring rest length']
-g = [0., 0., -9.81]
-
-# Try to read "air friction coefficient". If it does not exist C1 and C2 are
-# set to zero
-try:
-    C1 = settings['air friction coefficient'][0]
-    C2 = settings['air friction coefficient'][1]
-except KeyError:
-    C1 = 0.
-    C2 = 0.
-
-# variable order:
-# p1x p1y p1z p2x p2y p2z v1x v1y v1z v2x v2y v2z
-
-init_p1 = settings['initial position'][0]
-init_p2 = settings['initial position'][1]
-init_v1 = settings['initial velocity'][0]
-init_v2 = settings['initial velocity'][1]
-y0 = [init_p1[0], init_p1[1], init_p1[2], init_p2[0], init_p2[1], init_p2[2],
-      init_v1[0], init_v1[1], init_v1[2], init_v2[0], init_v2[1], init_v2[2]]
-t0 = 0.
-T = 2
+x1_0 = settings['initial position'][0]
+x2_0 = settings['initial position'][1]
+v1_0 = settings['initial velocity'][0]
+v2_0 = settings['initial velocity'][1]
 
 m = [settings['masses'][0], settings['masses'][1]]
 
+
+# See if we can also get friction coefficients. Otherwise set them to zero:
+try:
+    C1 = settings['air friction coefficient'][0]
+except KeyError:
+    C1 = 0.
+try:
+    C2 = settings['air friction coefficient'][1]
+except KeyError:
+    C2 = 0.
+
+
+# describe the differential equation as a first order ODE:
+y0 = [x1_0[0], x1_0[1], x1_0[2], x2_0[0], x2_0[1], x2_0[2],
[…]

### 2015-01-21 turcksin-spring-mechanics/paper.tex
Go through it one more time.

@@
 \eqref{eq:ode} divided by the $m_i$. With this, the code looks like this:
-\marginpar{After documenting spring.py, split it and simplify the first
-  version to the pieces we need here}
-\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]{spring.py}
+\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
+import sys
+import json
+import numpy
 
-\begin{wrapfigure}{R}{0.2\textwidth}
+# Read the name of the input file from the command line, and read options from the file:
+assert len(sys.argv) == 2, 'Please provide an input file.'
+with open(sys.argv[1], 'r') as f:
+    settings = json.loads(f.read())
+
+# Then retrieve the various parameters from what we just read:
+D = settings['spring constant']
+L = settings['spring rest length']
+x1_0 = settings['initial position'][0]
+x2_0 = settings['initial position'][1]
+v1_0 = settings['initial velocity'][0]
+v2_0 = settings['initial velocity'][1]
+
+m = [settings['masses'][0], settings['masses'][1]]
+
+# describe the differential equation as a first order ODE:
+y0 = [x1_0[0], x1_0[1], x1_0[2], x2_0[0], x2_0[1], x2_0[2],
+      v1_0[0], v1_0[1], v1_0[2], v2_0[0], v2_0[1], v2_0[2]]
+
+def f(t, y):
+    p1 = y[0:3]
+    p2 = y[3:6]
+    v1 = y[6:9]
+    v2 = y[9:12]
+
+    g = [0., 0., -9.81]
+
+    dist = numpy.linalg.norm(p2-p1)
+    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0]
+    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1]
+    return numpy.concatenate([v1, v2, a1, a2])
+
+# Next create an object that can integrate the ODE numerically:
+start_time = 0.
+end_time   = 5
+import scipy.integrate
+integrator = scipy.integrate.ode(f)
+integrator.set_integrator('vode', rtol=1e-6)
+integrator.set_initial_value(y0, start_time)
+
+# With this, do the integration step by step, appending values to an array in each step:
+t_values = [start_time]
+y_values = numpy.array([y0])
+while integrator.successful() and integrator.t < end_time:
+    integrator.integrate(end_time, step=True)
+    t_values.append(integrator.t)
+    y_values = numpy.vstack((y_values, integrator.y))
+
+# Having done so, output the number of time steps and the final positions:
+print "time steps:", len(t_values)
+print "final position:", y_values[len(t_values)-1,0:3], y_values[len(t_values)-1,3:6]
+\end{lstlisting}
+
+\begin{wrapfigure}{R}{0.3\textwidth}
   \begin{center}
     \vspace*{-24pt}
-    \includegraphics[width=0.18\textwidth]{tests/testcase-1.png}
+    \includegraphics[width=0.28\textwidth]{figures/testcase-1.png}
     \vspace*{-12pt}
@@
 canvas = fig.gca(projection='3d')
-canvas.plot(y_values[:,0], y_values[:,1], y_values[:,2], label='body 1')
-canvas.plot(y_values[:,3], y_values[:,4], y_values[:,5], label='body 2')
+canvas.plot(y_values[:,0], y_values[:,1], y_values[:,2],
+            label='body 1')
+canvas.plot(y_values[:,3], y_values[:,4], y_values[:,5], 
+            label='body 2')
 matplotlib.pyplot.show()
@@
 paper where they should be at time $t$. The following input file
-(\texttt{testcase-2.json}) does this:
-\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-2.json}
+(\texttt{tests/testcase-2.json}) does this:
+\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-2.json}
 
@@
 even though few people realize this when they start developing software. The
-issue essentially boils down to this: I need mechanisms that ensure that
-functionality if I have verified today works correctly, does not break when
-implementing new functionality. This is necessary because good programmers
+issue essentially boils down to this: We need mechanisms that ensure that
+the functionality we have verified works correctly today, does not break when
+we implement new functionality tomorrow. This is necessary because good programmers
 understand that they will make mistakes; what makes them good programmers is
@@
 true statement is that programs that are not tested produce wrong answers --
-where the emphasis is on the fact that they \textit{produce} wrong answers,
+where the emphasis is on the fact that they \textit{do} produce wrong answers,
 not that they \textit{may produce} wrong answers. Experienced programmers all
@@
 where exactly the problem is. On the other hand, if we had a testsuite that we
-run after essentially every change, we will be notified that something stopped
-working immediately -- with a relatively small number of changes made since
+run after essentially every change, we will be notified immediately that something stopped
+working -- with a relatively small number of changes made since
 the test last worked, and everything still fresh on our mind, finding the bug
@@
 \texttt{tests/} in your project and create two files in it, say
-\texttt{testcase-1.json} and \texttt{testcase-1.output}. The last part can
-be as easy as running commands such as
+\texttt{testcase-1.json} and the corresponding output
+\texttt{testcase-1.output}. When it comes time to test at a later time, step 3
+above can be as easy as running commands such as
 \begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
@@
 is now correct, then copy the new output over the new one. If you did not
-purposefully change it, then you know it's time to break out the debugger.
+purposefully change what the program computes, then you know it's time to break out the debugger.
 
@@
 use that the air friction force is proportional to the square of the speed and
-in the opposite direction of the velocity \cite{...}. Then our model
+in the opposite direction of the velocity. Then our model
 \eqref{eq:ode} needs to change to the following:
@@
 with the values in the input file if specified. A backward compatible choice
-of input parameter for the $C_i$ is to give them a default value of zero
+of input parameters for the $C_i$ is to give them a default value of zero
 because in that case equation \eqref{eq:ode2} equals the previous version,
@@
 
-Implementing all of this leads to the following program:
-\lstinputlisting[frame=single,basicstyle=\footnotesize,language=Python]{spring.py}
+Implementing all of this leads to the following extension of the program shown
+above%
+\footnote{This is the file \texttt{spring.py} in the directory corresponding to
+this paper}, with the new parts in lines 21--29 and 45--46:
+\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize,language=Python]{spring.py}
 
@@
 add to it something about air friction: 
-\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-3.json}
+\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-3.json}
 To verify that the results we get are correct, we can use that in this test
@@
 \texttt{friction[1]} in the code where we intended to use \texttt{friction[i]}.
-So let's copy \texttt{tests/testcase-3.json} to
-\texttt{tests/testcase-4.json}:
-\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-4.json}
+So let's copy \texttt{testcase-3.json} to
+\texttt{testcase-4.json}:
+\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-4.json}
 Here, the second ball has a greater mass, but the same air friction
@@
 Because it is so easy to create tests, it may be worthwhile cloning
-\texttt{tests/testcase-3.json} a second time into
-\texttt{tests/testcase-5.json} where now we just set the friction coefficients
+\texttt{testcase-3.json} a second time into
+\texttt{testcase-5.json} where now we just set the friction coefficients
 differently:
-\lstinputlisting[frame=single,basicstyle=\footnotesize]{tests/testcase-4.json}
+\lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize]{tests/testcase-5.json}
 Now the two balls have the same mass but the second one has a higher drag
@@
 behind the first.
+
+If we're reasonably convinced that the output is correct, then save the output
+in files \texttt{tests/testcase-3.output} through
+\texttt{tests/testcase-5.output} for use in future testing.
 
@@
 new test from scratch. To give an example of where this was necessary,
-consider \aspect{}: imagine that at the beginning, it could only compute
+consider \aspect{}: Imagine that at the beginning, it could only compute
 convection in a box. We would have testcases that described different
@@
 dot, is shown in Fig.~\ref{fig:aspect-tests} where we draw tests farther apart
-of there are more textual differences between the input files. Many tests come
+if there are more textual differences between the input files. Many tests come
 in groups of 2, 3 or 4 that only differ in a few lines of input -- simple
@@
 starting point to clone and graft a new testcase, you will not recall if
-\texttt{tests/testcase-3.json} or \texttt{tests/testcase-5.json} is the one
+\texttt{testcase-3.json} or \texttt{testcase-5.json} is the one
 that's closest to the situation you have in mind (for example, if you want to
@@
 such as the following will do just fine:
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]
-  # loop over all .json input files
-  for jsonfile in *.json ; do
-    # execute our program on it and pipe the output into a file
-    python ${jsonfile} > ${jsonfile}.new
-
-    # then compare output; only ask 'diff' to report whether the
-    # files are the same or not, and create output depending on this:
-    # print an 'X' if there is a difference, a '.' if there isn't. this
-    # ensures that test failures are easily visible
-    if (diff ${jasonfile} ${jsonfile}.new > /dev/null) ; then
-      echo " .    ${jasonfile}" ;
-    else 
-      echo " X    ${jasonfile}" ;
-    fi
-  done
-\end{lstlisting}
+\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=bash]{tests/compare.sh}
 If you call this script, say, \texttt{tests/compare.sh}, then running it using
@@
 to identify the simplest one of these (assuming that they were given useful
-names, see above in Section~\ref{sec:many-tests}), and that is the one one
+names, see above in Section~\ref{sec:many-tests}), and that is the one we
 will then use for debugging. If one breaks fundamental functionality somewhere
@@
 perspective, we have found that it is not usually a problem to have testcases
-that check overlapping parts of the software.
+that check overlapping parts of the software. Rather, we have come to the
+conclusion that having as many tests as possible, regardless of overlap.
 
@@
 will be executed in different order and the results will be different at the
-level of roundoff. In other words, your tests will start to fail even though
-nothing fundamentally changed -- but the output will be different.
+level of roundoff. In other words, your tests will fail even though
+nothing fundamentally changed.
 
@@
 
-A second area that creates problems that codes often try to compute zeros, but
-because of roundoff what is actually produced are numbers of the kind
+A second area that creates problems is that codes often try to compute zeros, but
+because of roundoff what is actually produced are numbers like
 4.863e-19. A human recognizes that this is a zero and, consequently, equal to
@@
 The solution to these problems that we have ultimately come to use
-is a program called \texttt{numdiff}. \texttt{numdiff} works essentially like
+is a program called \texttt{numdiff}.%
+\footnote{See \url{http://www.nongnu.org/numdiff/}.}
+\texttt{numdiff} works essentially like
 \texttt{diff} but it recognizes if the text it compares in two files is
@@
 
-When we talk to students who are just starting writing their own codes, then
+When we talk to students who just starting writing their own codes, then
 they typically understand the importance of testing in some abstract
@@
 a change might be. All experienced programmers will know this to be wrong. The
-reason is that over the course of a scientific code -- say, over the course of
+reason is that over the lifetime of a scientific code -- say, over the course of
 the 3 years a graduate student does research, and maybe even beyond this if
@@
 Foundation under award OCI-1148116 as part of the Software Infrastructure for
-Sustained Innovation (SI2) program; by the Computational
+Sustained Innovation (SI2) program; and by the Computational
 Infrastructure in Geodynamics initiative (CIG), through the National Science

### 2015-01-21 turcksin-spring-mechanics/tests/compare.sh

@@ -1,0 +1,15 @@
+# loop over all .json input files
+for jsonfile in *.json ; do
+  # execute our program on it and pipe the output into a file
+  python ${jsonfile} > ${jsonfile}.new
+
+  # then compare output; only ask 'diff' to report whether the
+  # files are the same or not, and create output depending on this:
+  # print an 'X' if there is a difference, a '.' if there isn't. this
+  # ensures that test failures are easily visible
+  if (diff ${jasonfile} ${jsonfile}.new > /dev/null) ; then
+    echo " .    ${jasonfile}" ;
+  else 
+    echo " X    ${jasonfile}" ;
+  fi
+done

### 2015-01-21 turcksin-spring-mechanics/tests/compare.sh
fix compare script

@@ -1,15 +1,15 @@
 # loop over all .json input files
 for jsonfile in *.json ; do
   # execute our program on it and pipe the output into a file
-  python ${jsonfile} > ${jsonfile}.new
+  python ../spring.py ${jsonfile} > ${jsonfile}.new
 
   # then compare output; only ask 'diff' to report whether the
   # files are the same or not, and create output depending on this:
   # print an 'X' if there is a difference, a '.' if there isn't. this
   # ensures that test failures are easily visible
-  if (diff ${jasonfile} ${jsonfile}.new > /dev/null) ; then
-    echo " .    ${jasonfile}" ;
+  if (diff ${jsonfile}.output ${jsonfile}.new > /dev/null) ; then
+    echo " .    ${jsonfile}" ;
   else 
-    echo " X    ${jasonfile}" ;
+    echo " X    ${jsonfile}" ;
   fi
 done

### 2015-01-21 turcksin-spring-mechanics/spring.py
simplify code
- move imports to top
- shorter assignments
- default values for missing friction
- add plotting code, add legend

@@ -1,8 +1,10 @@
 import sys
 import json
 import numpy
+import scipy.integrate
 
-# Read the name of the input file from the command line, and read options from the file:
+# Read the name of the input file from the command line, and read options from
+# the file:
 assert len(sys.argv) == 2, 'Please provide an input file.'
 with open(sys.argv[1], 'r') as f:
     settings = json.loads(f.read())
@@ -10,24 +12,12 @@ with open(sys.argv[1], 'r') as f:
 # Then retrieve the various parameters from what we just read:
 D = settings['spring constant']
 L = settings['spring rest length']
-x1_0 = settings['initial position'][0]
-x2_0 = settings['initial position'][1]
-v1_0 = settings['initial velocity'][0]
-v2_0 = settings['initial velocity'][1]
+m = settings['masses']
+[x1_0, x2_0] = settings['initial position']
+[v1_0, v2_0] = settings['initial velocity']
 
-m = [settings['masses'][0], settings['masses'][1]]
-
-
-# See if we can also get friction coefficients. Otherwise set them to zero:
-try:
-    C1 = settings['air friction coefficient'][0]
-except KeyError:
-    C1 = 0.
-try:
-    C2 = settings['air friction coefficient'][1]
-except KeyError:
-    C2 = 0.
-
+# the default friction is zero for both objects:
+[C1, C2] = settings.get('air friction coefficient', [0, 0])
 
 # describe the differential equation as a first order ODE:
 y0 = [x1_0[0], x1_0[1], x1_0[2], x2_0[0], x2_0[1], x2_0[2],
@@ -50,12 +40,12 @@ def f(t, y):
 # Next create an object that can integrate the ODE numerically:
 start_time = 0.
 end_time   = 5
-import scipy.integrate
 integrator = scipy.integrate.ode(f)
 integrator.set_integrator('vode', rtol=1e-6)
 integrator.set_initial_value(y0, start_time)
 
-# With this, do the integration step by step, appending values to an array in each step:
+# With this, do the integration step by step, appending values to an array in
+# each step:
 t_values = [start_time]
 y_values = numpy.array([y0])
 while integrator.successful() and integrator.t < end_time:
@@ -65,4 +55,17 @@ while integrator.successful() and integrator.t < end_time:
 
 # Having done so, output the number of time steps and the final positions:
 print "time steps:", len(t_values)
[…]

### 2015-01-21 turcksin-spring-mechanics/paper.tex
do not include code inside tex
instead put into subfolder and include. Update line numbers.

@@
 \eqref{eq:ode} divided by the $m_i$. With this, the code looks like this:
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
-import sys
-import json
-import numpy
 
-# Read the name of the input file from the command line, and read options from the file:
-assert len(sys.argv) == 2, 'Please provide an input file.'
-with open(sys.argv[1], 'r') as f:
-    settings = json.loads(f.read())
-
-# Then retrieve the various parameters from what we just read:
-D = settings['spring constant']
-L = settings['spring rest length']
-x1_0 = settings['initial position'][0]
-x2_0 = settings['initial position'][1]
-v1_0 = settings['initial velocity'][0]
-v2_0 = settings['initial velocity'][1]
-
-m = [settings['masses'][0], settings['masses'][1]]
-
-# describe the differential equation as a first order ODE:
-y0 = [x1_0[0], x1_0[1], x1_0[2], x2_0[0], x2_0[1], x2_0[2],
-      v1_0[0], v1_0[1], v1_0[2], v2_0[0], v2_0[1], v2_0[2]]
-
-def f(t, y):
-    p1 = y[0:3]
-    p2 = y[3:6]
-    v1 = y[6:9]
-    v2 = y[9:12]
-
-    g = [0., 0., -9.81]
-
-    dist = numpy.linalg.norm(p2-p1)
-    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0]
-    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1]
-    return numpy.concatenate([v1, v2, a1, a2])
-
-# Next create an object that can integrate the ODE numerically:
-start_time = 0.
-end_time   = 5
-import scipy.integrate
-integrator = scipy.integrate.ode(f)
-integrator.set_integrator('vode', rtol=1e-6)
-integrator.set_initial_value(y0, start_time)
-
-# With this, do the integration step by step, appending values to an array in each step:
-t_values = [start_time]
-y_values = numpy.array([y0])
-while integrator.successful() and integrator.t < end_time:
-    integrator.integrate(end_time, step=True)
-    t_values.append(integrator.t)
-    y_values = numpy.vstack((y_values, integrator.y))
-
-# Having done so, output the number of time steps and the final positions:
-print "time steps:", len(t_values)
-print "final position:", y_values[len(t_values)-1,0:3], y_values[len(t_values)-1,3:6]
-\end{lstlisting}
+\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
+{codeinclude/spring1.py}
 
@@
 get that by adding the following lines to the bottom of the script:
-\begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
-import matplotlib.pyplot
-from mpl_toolkits.mplot3d import Axes3D
-fig = matplotlib.pyplot.figure()
-canvas = fig.gca(projection='3d')
-canvas.plot(y_values[:,0], y_values[:,1], y_values[:,2],
-            label='body 1')
-canvas.plot(y_values[:,3], y_values[:,4], y_values[:,5], 
-            label='body 2')
-matplotlib.pyplot.show()
-\end{lstlisting}
+\lstinputlisting[frame=single,basicstyle=\footnotesize,numbers=left,language=Python]
+{codeinclude/plot.py}
 The output this creates is shown in Fig.~\ref{fig:testcase-1}.
@@
 \footnote{This is the file \texttt{spring.py} in the directory corresponding to
-this paper}, with the new parts in lines 21--29 and 45--46:
+this paper}, with the new parts in lines 19--20 and 35--36:
 \lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize,language=Python]{spring.py}

### 2015-01-21 turcksin-spring-mechanics/codeinclude/plot.py

@@ -1,0 +1,10 @@
+import matplotlib.pyplot
+from mpl_toolkits.mplot3d import Axes3D
+fig = matplotlib.pyplot.figure()
+canvas = fig.gca(projection='3d')
+canvas.plot(y_values[:,0], y_values[:,1], y_values[:,2],
+            label='body 1')
+canvas.plot(y_values[:,3], y_values[:,4], y_values[:,5], 
+            label='body 2')
+canvas.legend()
+matplotlib.pyplot.show()

### 2015-01-21 turcksin-spring-mechanics/codeinclude/spring1.py

@@ -1,0 +1,55 @@
+import sys
+import json
+import numpy
+import scipy.integrate
+
+# Read the name of the input file from the command line, and read options from
+# the file:
+assert len(sys.argv) == 2, 'Please provide an input file.'
+with open(sys.argv[1], 'r') as f:
+    settings = json.loads(f.read())
+
+# Then retrieve the various parameters from what we just read:
+D = settings['spring constant']
+L = settings['spring rest length']
+m = settings['masses']
+[x1_0, x2_0] = settings['initial position']
+[v1_0, v2_0] = settings['initial velocity']
+
+# describe the differential equation as a first order ODE:
+y0 = [x1_0[0], x1_0[1], x1_0[2], x2_0[0], x2_0[1], x2_0[2],
+      v1_0[0], v1_0[1], v1_0[2], v2_0[0], v2_0[1], v2_0[2]]
+
+def f(t, y):
+    p1 = y[0:3]
+    p2 = y[3:6]
+    v1 = y[6:9]
+    v2 = y[9:12]
+
+    g = [0., 0., -9.81]
+
+    dist = numpy.linalg.norm(p2-p1)
+    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0] - C1*numpy.linalg.norm(p1)*p1
+    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1] - C2*numpy.linalg.norm(p2)*p2
+    return numpy.concatenate([v1, v2, a1, a2])
+
+
+# Next create an object that can integrate the ODE numerically:
+start_time = 0.
+end_time   = 5
+integrator = scipy.integrate.ode(f)
+integrator.set_integrator('vode', rtol=1e-6)
+integrator.set_initial_value(y0, start_time)
+
+# With this, do the integration step by step, appending values to an array in
+# each step:
+t_values = [start_time]
+y_values = numpy.array([y0])
+while integrator.successful() and integrator.t < end_time:
+    integrator.integrate(end_time, step=True)
+    t_values.append(integrator.t)
+    y_values = numpy.vstack((y_values, integrator.y))
+
+# Having done so, output the number of time steps and the final positions:
+print "time steps:", len(t_values)
+print "final position:", y_values[-1,0:3], y_values[-1,3:6]

### 2015-01-22 turcksin-spring-mechanics/tests/compare.sh
change filenames for output/reference

@@ -1,15 +1,16 @@
 # loop over all .json input files
 for jsonfile in *.json ; do
   # execute our program on it and pipe the output into a file
-  python ../spring.py ${jsonfile} > ${jsonfile}.new
+  testname=`basename $jsonfile .json`
+  python ../spring.py ${jsonfile} > ${testname}.output
 
   # then compare output; only ask 'diff' to report whether the
   # files are the same or not, and create output depending on this:
   # print an 'X' if there is a difference, a '.' if there isn't. this
   # ensures that test failures are easily visible
-  if (diff ${jsonfile}.output ${jsonfile}.new > /dev/null) ; then
+  if (diff ${testname}.output ${testname}.reference > /dev/null) ; then
     echo " .    ${jsonfile}" ;
-  else 
+  else
     echo " X    ${jsonfile}" ;
   fi
 done

### 2015-01-22 turcksin-spring-mechanics/paper.tex
correct shell commands in paper

@@
 \texttt{testcase-1.json} and the corresponding output
-\texttt{testcase-1.output}. When it comes time to test at a later time, step 3
+\texttt{testcase-1.reference}. When it comes time to test at a later time, step 3
 above can be as easy as running commands such as
 \begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
-  python masses-and-springs.py testcase-1.json > testcase-1.output.new
-  diff testcase-1.output testcase-1.output.new
+  python spring.py testcase-1.json > testcase-1.output
+  diff testcase-1.output testcase-1.reference
 \end{lstlisting}
@@
 same and your program computed the same answer on this input file as it did
-back when you created the \texttt{testcase-1.output} file. If there is a
+back when you created the \texttt{testcase-1.reference} file. If there is a
 difference, then you know that you just broke one of the testcases. If you
@@
 \begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
-  python masses-and-springs.py testcase-1.json > testcase-1.output.new
-  diff testcase-1.output testcase-1.output.new
-  python masses-and-springs.py testcase-2.json > testcase-2.output.new
-  diff testcase-2.output testcase-2.output.new
+  python spring.py testcase-1.json > testcase-1.output
+  diff testcase-1.output testcase-1.reference
+  python spring.py testcase-2.json > testcase-2.output
+  diff testcase-2.output testcase-2.reference
 \end{lstlisting}
@@
 \begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
-  numdiff -a 1e-6 -r 1e-8 testcase-1.output testcase-1.output.new
+  numdiff -a 1e-6 -r 1e-8 testcase-1.output testcase-1.reference
 \end{lstlisting}
 Using this trick, numbers that are written into the output file
-\texttt{testcase-1.output.new} on one machine with slight differences to the
-ones originally saved in \texttt{testcase-1.output} will still be considered
+\texttt{testcase-1.output} on one machine with slight differences to the
+ones originally saved in \texttt{testcase-1.reference} will still be considered
 the same, and the comparison will succeed.

### 2015-01-22 turcksin-spring-mechanics/paper.tex
fix typos and small edits

@@
    Timo Heister\footnote{Mathematical Sciences, O-110 Martin Hall, Clemson University,
-     Clemson, SC 29634-095, USA, \url{heister@clemson.edu}}
+     Clemson, SC 29634-0975, USA, \url{heister@clemson.edu}}
 \and
@@
 
-\section{\aspect{}: the Advanced Solver for Problems in Earth ConvecTion}
+\section{\aspect{}: the Advanced Solver for Problems in Earth's ConvecTion}
 \label{sec:aspect}
@@
 The code that made us think about how to test whole applications is called
-\aspect{} -- short for the Advanced Solver for Problems in Earth ConvecTion
+\aspect{} -- short for the Advanced Solver for Problems in Earth's ConvecTion
 \cite{KHB12,aspectmanual,aspectweb}. It is a program that simulates how material moves around the Earth
@@
 linking to the videos at \url{http://www.youtube.com/embed/j63MkEc0RRw} and
-\url{http://www.youtube.com/embed/EJJ6f4hmDPU}, of only because they are pretty.
+\url{http://www.youtube.com/embed/EJJ6f4hmDPU}, if only because they are pretty.
 
@@
 \footnote{In order to run this program, you will have to have the Python
-  NumPy, SciPy, and MatPlotLib packages installed on your system, along with
+  NumPy, SciPy, and matplotlib packages installed on your system, along with
   the basic Python interpreter.}
 The program we'll show you in a second reads its input parameters from a file in JSON format%
-\footnote{JSON is not the prettiest format to write input files in and we
+\footnote{While certainly more sane compared to XML, JSON is not the prettiest format to write input files in and we
   don't recommend you use it for your programs (it is typically intended for
@@
 y(0)=\mathbf y_0$, and tell the integrator about $\mathbf f$ and $\mathbf
-y_0$. We do this by introducing variables $\mathbf v_i=\mathbf x'_i$ and then
+y_0$. We do this by introducing variables for the velocities $\mathbf v_i=\mathbf x'_i$ and then
 setting $\mathbf
@@
 \footnote{This is the file \texttt{spring.py} in the directory corresponding to
-this paper}, with the new parts in lines 19--20 and 35--36:
+this paper.}, with the new parts in lines 19--20 and 35--36:
 \lstinputlisting[frame=single,numbers=left,basicstyle=\footnotesize,language=Python]{spring.py}
@@
 that check overlapping parts of the software. Rather, we have come to the
-conclusion that having as many tests as possible, regardless of overlap.
+conclusion that it is best to have as many tests as possible, regardless of overlap.
 
@@
 Davis.
-\marginpar{Timo -- please update.}
+
+T.~Heister was partially supported by the Computational
+Infrastructure in Geodynamics initiative (CIG), through the National Science
+Foundation under Award No.~EAR-0949446 and The University of California --
+Davis.
 

### 2015-01-22 turcksin-spring-mechanics/paper.tex
Fix a bug in the code and a typo in the paper.

@@
 \author{Bruno Turcksin\footnote{Department of Mathematics, Texas A\&M
-    University, College Station, TX 77843-3368, USA, \url{???}}
+    University, College Station, TX 77843-3368, USA, \url{turcksin@math.tamu.edu}}
 \and
@@
 
-\marginpar{Bruno - please update email address}
 \section{Introduction}
@@
 
-When we talk to students who just starting writing their own codes, then
+When we talk to students who just start writing their own codes, then
 they typically understand the importance of testing in some abstract

### 2015-01-22 turcksin-spring-mechanics/codeinclude/spring1.py

@@ -29,8 +29,8 @@ def f(t, y):
     g = [0., 0., -9.81]
 
     dist = numpy.linalg.norm(p2-p1)
-    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0] - C1*numpy.linalg.norm(p1)*p1
-    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1] - C2*numpy.linalg.norm(p2)*p2
+    a1 = g - D*(dist-L) * (p1-p2)/dist/m[0]
+    a2 = g - D*(dist-L) * (p2-p1)/dist/m[1]
     return numpy.concatenate([v1, v2, a1, a2])
 
 

### 2015-01-23 turcksin-spring-mechanics/paper.tex
add two points

@@
 Developing testing schemes for libraries is reasonably well understood. So,
-when we ventured in the area of building applications for scientific purposes
+when we ventured in the area of building \textit{applications} for scientific purposes
 on top of the libraries we had created, we had initially thought that one can
@@
 call a function in a library by presenting a magic \texttt{main()} function in
-a unit test.
+a unit test. There are of course more advanced ways of testing features of an
+application in isolation by replacing other parts of the code during testing by so-called
+mock or stub objects. We will not discuss this here though.
 
@@
   the basic Python interpreter.}
+Note that testing is even more important in dynamically typed languages like
+Python, because you do not have the compiler to help you find bugs at compile 
+time.
+
 The program we'll show you in a second reads its input parameters from a file in JSON format%

### 2015-01-25 turcksin-spring-mechanics/tests/compare.sh
Revert order of 'diff' commands; a couple of additions.

@@ -8,7 +8,7 @@ for jsonfile in *.json ; do
   # files are the same or not, and create output depending on this:
   # print an 'X' if there is a difference, a '.' if there isn't. this
   # ensures that test failures are easily visible
-  if (diff ${testname}.output ${testname}.reference > /dev/null) ; then
+  if (diff ${testname}.reference ${testname}.output > /dev/null) ; then
     echo " .    ${jsonfile}" ;
   else
     echo " X    ${jsonfile}" ;

### 2015-01-31 turcksin-spring-mechanics/paper.tex

@@
   python spring.py testcase-1.json > testcase-1.output
-  diff testcase-1.output testcase-1.reference
+  diff testcase-1.reference testcase-1.output
 \end{lstlisting}
@@
 or because you fixed behavior you know was erroneous and have verified that it
-is now correct, then copy the new output over the new one. If you did not
+is now correct, then copy the new output over the old reference file. If you did not
 purposefully change what the program computes, then you know it's time to break out the debugger.
@@
   python spring.py testcase-1.json > testcase-1.output
-  diff testcase-1.output testcase-1.reference
+  diff testcase-1.reference testcase-1.output
   python spring.py testcase-2.json > testcase-2.output
-  diff testcase-2.output testcase-2.reference
+  diff testcase-2.reference testcase-2.output
 \end{lstlisting}
@@
 If we're reasonably convinced that the output is correct, then save the output
-in files \texttt{tests/testcase-3.output} through
-\texttt{tests/testcase-5.output} for use in future testing.
+in files \texttt{tests/testcase-3.reference} through
+\texttt{tests/testcase-5.reference} for use in future testing.
 
@@
 \begin{lstlisting}[frame=single,basicstyle=\footnotesize,numbers=left,language=csh]
-  numdiff -a 1e-6 -r 1e-8 testcase-1.output testcase-1.reference
+  numdiff -a 1e-6 -r 1e-8 testcase-1.reference testcase-1.output
 \end{lstlisting}
